CRUSE to Safe Cycling in Ireland \\\large{An Open Source Methodology to Support Active Travel}

### 2023-09-20 paper.qmd
Add parkin2018 citation
Updates for .tex version

# # Uncomment for formatted version
# format:
#   elsevier-pdf:
#     keep-tex: true
#     journal:
#       name: Journal Name
#       formatting: preprint
#       model: 3p
#       cite-style: super

@@

A feature of active travel interventions is that they require dense networks of routes to be effective \[TODO: cite parkin\].
This means that high levels of geographic resolution are needed in the data used to estimate cycling potential.
However, datasets on travel patterns are often only available at the level of administrative zones.
The Central Statistics Office (CSO) in Ireland provides Place of Work, School or College Census of Anonymised Records ([POWSCAR](https://www.cso.ie/en/census/census2016reports/powscar/)) data on the number of people travelling to work and school at the Electoral Division (ED) level, for example.

@@

![](images/paste-1.png)

>>>>

# Uncomment for formatted version
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: super

@@

A feature of active travel interventions is that they require dense networks of routes to be effective [@parkin2018].
This means that high levels of geographic resolution are needed in the data used to estimate cycling potential.
However, datasets on travel patterns are often only available at the level of administrative zones.
The Central Statistics Office (CSO) in Ireland provides Place of Work, School or College Census of Anonymised Records ([POWSCAR](https://www.cso.ie/en/census/census2016reports/powscar/)) data on the number of people travelling to work and school at the Electoral Division (ED) level, for example.

@@

::: {#fig-dublin}
![](images/paste-1.png)

![](images/paste-3.png)

Illustration of the density of route networks generated by the CRUSE tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland.
Source: CRUSE tool, available at <https://cruse.bike/>.
:::

### 2023-09-20 paper.tex
Updates for .tex version

\begin{keyword}
    Cycling \sep Open-source \sep Road Safety \sep 
    Active Travel
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

Transport is a major contributor to climate change, premature death due
to road traffic collisions and air pollution, yet has the potential to
be a major contributor to decarbonization and improved public health.
Transport is responsible for 23\% of global emissions, 70\% of which is
from road transport, nearly half of which is from passenger cars
\citep{jaramillo2022}. The transport system encourages, enables and in
some cases enforces unsustainable lifestyles, including over-consumption
of goods due to excessive mobile storage space and dependency on
services that are only accessible by car due to land use plans that have
built up around roads \citep{gray2001, shergold2012, motte-baumvol2010}.

Recognizing the impacts of poorly designed and performing transport
systems on their citizens, governments in many countries have set
targets and taken actions and set targets. In the context of climate,
road safety and physicial inactivity crises, policies to improve
transport systems can be classified according to the
`Avoid-Shift-Improve' (ASI) framework \citep{jaramillo2022}. The
framework highlights the importance of demand reduction (\emph{avoid}ing
unnecessary trips), in addition to mode \emph{shift} and
\emph{improvement} of existing energy converters, in that order.

Uptake of cycling, the main topic of this paper, should be seen in this
broader context of transport decarbonisation \citep{brand2020}. Although
cycling uptake appears on the surface to only relate to the `shift' part
of the ASI framework, closer consideration of the knock-on impacts of
cycling uptake shows that it can also help avoid unnecessary trips
{[}citation needed, any ideas Hussein?{]} . The rapid uptake of highly
efficient e-bikes can also be seen as an improvement on most electric
vehicles, which are too heavy and expensive to be a sustainable
alternative and could in fact delay the transition away from car
dependency {[}citation needed{]}.

At the European Union level \ldots{}

In Ireland\ldots{}

Mobility hub.

Business cases need evidence.

Como UK.

{[}Gary, Suzanne, others input here please.{]}

In this context, TII commissioned the CRUSE project which is\ldots{}

The rest of paper is structured as follows\ldots{}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

{[}Should this bit go in a subsection in the Introduction rather than in
the methods? (RL){]}

The methods used to generate the evidence presented in the CRUSE tool
for Ireland build on previous work, particularly the Propensity to Cycle
Tool (PCT), which was originally developed for the UK's Department for
Transport and which is publicly available at www.pct.bike
\citep{lovelace2017}. The PCT approach has had major policy impacts, as
outlined in Research Excellence Framework (REF) impact case studies
submitted by the University of Leeds and the University of Westminster,
which demonstrate that the tool ``revolutionised strategic cycle
planning in England and Wales''\footnote{See REF Impact Case Study
  ``Cycle network policy, planning and investment transformed by the
  Propensity to Cycle Tool'' at
  \href{https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f}{results2021.ref.ac.uk}
  submitted by the University of Leeds.} by overcoming the barriers to
cycling investment imposed by lack of evidence on cycling
potential\footnote{See the REF Impact Case Study ``Creating Step Changes
  in Cycling Policy and Infrastructure Planning across the UK'' at
  \href{https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B}{results2021.ref.ac.uk}
  submitted by the University of Westminster.} .

The first version of the PCT was based solely on current and future
potential uptake of cycling for single stage travel to work at desire
line, zone, route, and route network levels \citep{lovelace2016}. It was
launched in April 2017 as the government-endorsed tool for strategic
cycle network planning, as part of the Cycling and Walking Investment
Strategy \citep{cycling2017}. Extensions of the PCT approach have
included estimation of benefits at the individual level
\citep{woodcock2018}, addition of travel to school network
\citep{goodman2019}, and improvement modelling of impacts on health,
environmental and distributional outcomes \citep{woodcock2021}. In
Portugal, a cycling uptake tool to support decision making was developed
based on PCT methods, for the Lisbon metro region (see
\href{https://biclar.tmlmobilidade.pt}{biclaR}, and again, social and
environment impacts were added as a relevant information to support
decisions, using the WHO HEAT for Cycling tool\footnote{See the Health
  economic assessment tool for walking and cycling at
  \href{https://www.heatwalkingcycling.org/\#homepage}{www.heatwalkingcycling.org}.}
\citep{felix2023}. biclaR added on the previous methods by setting an
`intermodality' scenario that combines cycling with currently available
public transit options based on General Transit Feed Specification
(GTFS) data.

Bit on Jittering \citep{lovelace2022}

Despite these advances, the PCT approach outlined in previous research
paper had limitations:

\begin{itemize}
\item
  Limited coverage of modes beyond travel to work and school
\item
  \ldots{}
\end{itemize}

\hypertarget{results}{%
\section{Results}\label{results}}

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

@@

>>>>

\begin{keyword}
    Cycling \sep Open-source \sep Road Safety \sep 
    Active Travel
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, frame hidden, sharp corners, enhanced]}{\end{tcolorbox}}\fi

@@

Mobility kills. High energy transport systems are a major contributor to
climate change, a leading cause of premature death and injury due to
road traffic collisions, and a cause of disease due to airborne
pollutants. Conversely, evidence-based and effective transport
\emph{policies} have great potential to decarbonize economies, improve
public health, and save lives.

Transport is responsible for 23\% of global emissions, 70\% of which is
from road transport, nearly half of which (around 10\% of global
emissions) is from passenger cars \citep{jaramillo2022}. The transport
system encourages, enables and in some cases enforces unsustainable
lifestyles, including over-consumption of goods due to excessive mobile
storage space and dependency on services that are only accessible by car
due to land use plans that have built up around roads
\citep{gray2001, shergold2012, motte-baumvol2010}.

Recognizing the growing evidence of such impacts of poorly designed and
performing transport systems, governments in many countries have set
targets and taken actions. In the context of climate, road safety and
physicial inactivity crises, policies to improve transport systems can
be classified according to the `Avoid-Shift-Improve' (ASI) framework
\citep{jaramillo2022}. The framework highlights the importance of demand
reduction (\emph{avoid}ing unnecessary trips), in addition to mode
\emph{shift} and \emph{improvement} of existing energy converters, in
that order.

Uptake of cycling, the main topic of this paper, should be seen in this
broader context of transport decarbonisation \citep{brand2020} and
sustainable mobility \citep{burns2013}. Although cycling uptake appears
on the surface to only relate to the `shift' part of the ASI framework,
closer consideration of the knock-on impacts of cycling uptake shows
that it can also help avoid unnecessary trips \citep{nello-deakin2020}.
The rapid uptake of highly efficient e-bikes can also be seen as an
improvement on most electric vehicles, which are too heavy and expensive
to be a sustainable alternative and could in fact delay the transition
away from car dependency and inadvertently enable ``high travel
lock-in'' \citep{anable2019}. At the European level, the European Union
has a target of reducing greenhouse gas emissions by 55\% by 2030,
compared to 1990 levels, and to achieve `net-zero' by 2050.{[}\^{}1{]}

Regarding road traffic casualties, another deadly consequence of
inefficient transport systems, the Road Infrastructure Safety Management
(RISM) directive (2008/96/EC) requires member states to implement a road
safety management system (RSMS) for all public roads. Specifically,
``Member States shall ensure that the ranking of high accident
concentration sections and the network safety ranking are carried
out''.\footnote{https://eur-lex.europa.eu/legal-content/EN/TXT/?u
  ri=celex\%3A32008L0096} Given that `safety' in this context is
usefully quantified as the number of people killed and seriously injured
(KSI) per distance travelled, the directive requires estimation of
distance travelled by mode, down to the road link level. For active
modes, about which there is a paucity of data compared with motorized
modes, this is a major challenge. Better data to inform road safety
policies and interventions is a motivation for the CRUSE tool outlined
in this paper, and modeling active travel more broadly.

In Ireland, the Road Safety Authority (RSA) has set the target of
halving the number of road traffic deaths and serious injuries by
2030.{[}\^{}3{]} Doing so while simultaneously enabling rapid uptake of
active modes will require key active travel routes to be identified and
improved.

It was in this context that Transport Infrastructure Ireland (TII)
commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE)
project. Based on the Propensity to Cycle Tool (PCT) for England and
Wales \citep{lovelace2017}, the CRUSE tool was developed to provide
evidence on current cycling levels and future cycling potential
nationwide across Ireland.

The rest of paper is structured as follows. In
Section~\ref{sec-methods}, we outline the methods used to generate the
evidence presented in the CRUSE tool for Ireland. In
Section~\ref{sec-results}, we present the results of the CRUSE tool,
including estimates of current cycling levels and future cycling
potential at the national, regional and local levels. In
Section~\ref{sec-discussion}, we discuss the implications of the results
for policy and practice. In Section~\ref{sec-conclusions}, we conclude
the paper.

\hypertarget{sec-methods}{%
\section{Methods and data}\label{sec-methods}}

The methods used to generate the evidence presented in the CRUSE tool
for Ireland build on the Propensity to Cycle Tool (PCT), which was
originally funded by the UK's Department for Transport and developed by
a multi-university team. An important feature of the PCT is that it is
open source and publicly available (at
\href{https://www.pct.bike/}{www.pct.bike}), allowing its use by all
stakeholders in the transport planning process \citep{lovelace2017}. The
PCT approach has had major policy impacts, as outlined in Research
Excellence Framework (REF) impact case studies, which demonstrate that
the tool ``revolutionised strategic cycle planning in England and
Wales''\footnote{See REF Impact Case Study ``Cycle network policy,
  planning and investment transformed by the Propensity to Cycle Tool''
  at
  \href{https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f}{results2021.ref.ac.uk}
  submitted by the University of Leeds.} by overcoming the barriers to
cycling investment imposed by lack of evidence on cycling
potential\footnote{See the REF Impact Case Study ``Creating Step Changes
  in Cycling Policy and Infrastructure Planning across the UK'' at
  \href{https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B}{results2021.ref.ac.uk}
  submitted by the University of Westminster.} .

The first version of the PCT was based on current and future potential
uptake of cycling for single stage \emph{travel to work} at desire line,
zone, route, and route network levels \citep{lovelace2016}. It was
launched in April 2017 as the government-endorsed tool for strategic
cycle network planning, as part of the Cycling and Walking Investment
Strategy \citep{cycling2017}. Extensions of the PCT approach have
included estimation of benefits at the individual level
\citep{woodcock2018}, addition of travel to school network
\citep{goodman2019}, and improved modelling of impacts on health,
environmental and distributional outcomes \citep{woodcock2021}.
Initially developed just for England, the PCT was extended to cover all
of Wales (for commuter data only) in 2018.

The PCT approach has been applied in other countries, including Ireland
(the topic of this paper), Scotland, and Portugal. In Portugal the
`biclaR' project, based on methods underlying the PCT, has been
developed and deployed for the Lisbon metro region. The resulting
evidence is available in an interactive web application hosted at
\href{https://biclar.tmlmobilidade.pt}{biclar.tmlmobilidade.pt}
\citep{felix2023}. biclaR includes estimates of impacts, using the World
Health Organisation (WHO) `HEAT for Cycling' tool\footnote{See the
  Health economic assessment tool for walking and cycling at
  \href{https://www.heatwalkingcycling.org/\#homepage}{www.heatwalkingcycling.org}.}
and an `intermodality' scenario that combines cycling with currently
available public transit options based on General Transit Feed
Specification (GTFS) data.

The CRUSE tool seeks to overcome the following methodological
limitations of the original PCT:

\begin{itemize}
\tightlist
\item
  Low resolution of data, with routes starting and ending in
  administrative zone centroids
\item
  Limited coverage of trip purposes beyond travel to work and school
\item
  A web interface that was not user-friendly or intuitive
\end{itemize}

Methods were developed to overcome each of these, as outlined in
Section~\ref{sec-trip-purposes} to Section~\ref{sec-ui}.

\hypertarget{sec-disaggregation}{%
\subsection{Disaggregation of origin-destination
data}\label{sec-disaggregation}}

A feature of active travel interventions is that they require dense
networks of routes to be effective \citep{parkin2018}. This means that
high levels of geographic resolution are needed in the data used to
estimate cycling potential. However, datasets on travel patterns are
often only available at the level of administrative zones. The Central
Statistics Office (CSO) in Ireland provides Place of Work, School or
College Census of Anonymised Records
(\href{https://www.cso.ie/en/census/census2016reports/powscar/}{POWSCAR})
data on the number of people travelling to work and school at the
Electoral Division (ED) level, for example.

The method used to convert OD data to route networks used in the PCT was
to calculate a single route between the population weighted centroids
associated with each OD pair. This method works fine when the OD data
represents movement between small areas, but was not appropriate for
generating route networks from the POWSCAR data because zone centroids
are so far apart that the resulting route networks would be sparse and
unrealistic. To tackle this issue we developed a new method for OD data
disaggregation called `jittering' \citep{lovelace2022}. The method works
by first disaggregating the OD data based on a `disaggregation
threshold' and then assigning each disaggregated `sub-OD' pair to
`subpoints' within each zone. As shown in Figure~\ref{fig-dublin}, the
resulting route networks are dense, even in rural areas.

\begin{figure}

{\centering

\includegraphics{images/paste-1.png}

\includegraphics{images/paste-3.png}

}

\caption{\label{fig-dublin}Illustration of the density of route networks
generated by the CRUSE tool for Dublin City and surroundings (top) and
rural County Mayo on the West coast of Ireland. Source: CRUSE tool,
available at \url{https://cruse.bike/}.}

\end{figure}

\hypertarget{sec-trip-purposes}{%
\subsection{Additional trip purposes}\label{sec-trip-purposes}}

bla bla

\hypertarget{sec-ui}{%
\subsection{User interface}\label{sec-ui}}

\hypertarget{sec-results}{%
\section{Results}\label{sec-results}}

@@

\hypertarget{sec-discussion}{%
\section{Discussion}\label{sec-discussion}}

\hypertarget{sec-conclusions}{%
\section{Conclusions}\label{sec-conclusions}}

@@

CSO: Central Statistics Office

CRUSE: Cycle Route Uptake and Scenario Estimation

ED: Electoral Division

GTFS: General Transit Feed Specification

KSI: Killed and Seriously Injured

NRA: National Road Authority

NRN: National Road Network

OD: Origin-Destination, typically referring to origin-destination data
which contains information on the number of people travelling between
each pair of zones

OSM: Open Street Map

PCT: Propensity to Cycle Tool

POWSCAR: Place of Work, School or College Census of Anonymised Records

TII: Transport Infrastructure Ireland
